---
title: "Differential Gene Expression Analysis - Delta8-THC Vaping Study"
author: "Martha Pierson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---
```{r setup, include=FALSE}
# Save snakemake object for interactive debugging
# snakemake <- readRDS(".deg.Rmd.RDS")
saveRDS(snakemake, ".deg.Rmd.RDS")

# Load all required libraries
library(edgeR)        # RNA-seq differential expression
library(DESeq2)       # normalization and DEG analysis
library(limma)        # linear models for DEG analysis
library(data.table)   # fast data manipulation
library(ggplot2)      # plotting
library(reshape2)     # data reshaping for plots
library(pheatmap)     # heatmaps
library(RColorBrewer) # color palettes
library(dplyr)        # data manipulation
library(tibble)       # modern data frames
library(fgsea)        # gene set enrichment analysis
library(msigdbr)      # MSigDB gene sets
```

# Load gene count matrices
```{r load_counts, include=TRUE}
# Read all featureCounts output files listed in snakemake input
# Each featureCounts file corresponds to one sample (56 in our case)
files <- snakemake@input$counts

# readDGE reads multiple featureCounts files and merges them into one matrix
# columns 1 and 7 are the gene ID and count columns in featureCounts output
# comment.char="#" skips the header lines that start with #
y <- readDGE(files, columns = c(1, 7), comment.char = "#")
count_matrix <- y$counts

# Rename columns from full file paths to just sample IDs
# e.g. "/path/to/SRR37298734.featureCounts.txt" -> "SRR37298734"
colnames(count_matrix) <- gsub(".featureCounts.txt", "", basename(files))

# Preview the count matrix
head(count_matrix)
```

# Load samplesheet
```{r load_samplesheet, include=TRUE}
# Read sample metadata file
samplesheet <- fread(snakemake@input$samplesheet)

# Make sure sample_id column matches count matrix column names
samplesheet <- samplesheet[match(colnames(count_matrix), samplesheet$sample_id), ]

# Convert product_type to a factor so limma treats it as categorical
# Setting control as the reference level means all comparisons are made against it
samplesheet$product_type <- relevel(factor(samplesheet$product_type), ref = "control")

# Convert donor to a factor (categorical variable) for the linear model
samplesheet$donor <- factor(samplesheet$donor)

# Preview the samplesheet
head(samplesheet)
```

# Filter lowly-expressed genes
```{r filter_counts, include=TRUE}
# Create a DGEList object - this is limma/edgeR's core data structure
# that stores counts and sample information together
y <- DGEList(counts = count_matrix, samples = samplesheet)

# Calculate counts per million (CPM) for filtering
# CPM normalizes for library size so we can compare across samples
cpm_matrix <- cpm(y)

# Keep genes that have at least 1 CPM in at least 4 samples
# (4 = smallest group size, i.e. one donor per treatment)
# This removes genes that are too lowly expressed to be meaningful
keep <- rowSums(cpm_matrix >= 1) >= 4
y_filt <- y[keep, ]

# Report how many genes were kept vs removed
cat("Genes before filtering:", nrow(y), "\n")
cat("Genes after filtering:", nrow(y_filt), "\n")
cat("Genes removed:", nrow(y) - nrow(y_filt), "\n")
```

# Normalize counts
```{r normalize, include=TRUE}
# Calculate normalization factors using TMM (Trimmed Mean of M-values)
# TMM accounts for differences in library composition between samples
# i.e. if one sample has a few highly expressed genes dominating the counts,
# TMM corrects for this so samples are comparable
y_filt <- calcNormFactors(y_filt, method = "TMM")

# Check the normalization factors - values close to 1 are good
# Values very different from 1 suggest unusual samples
y_filt$samples$norm.factors
```

# Boxplot of log2 CPM per sample
```{r boxplot_raw, include=TRUE, fig.width=12, fig.height=6}
# Convert counts to log2 CPM for visualization
# Adding 1 before log transformation avoids log(0) which is undefined
logcpm <- cpm(y_filt, log = TRUE, prior.count = 1)

# Melt the matrix from wide to long format for ggplot
melted <- reshape2::melt(logcpm, varnames = c("Gene", "Sample"),
                          value.name = "logCPM")

# Add treatment information for coloring
melted$product_type <- samplesheet$product_type[match(melted$Sample,
                                                       samplesheet$sample_id)]

# Boxplot of log2 CPM per sample
ggplot(melted, aes(x = Sample, y = logCPM, fill = product_type)) +
    geom_boxplot(outlier.size = 0.2) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(title = "Log2 CPM distribution per sample",
         x = "Sample",
         y = "Log2 CPM",
         fill = "Product Type")
```

# Set up limma-voom model
```{r limma_setup, include=TRUE}
# Create the design matrix - this defines our linear model
# ~ donor + product_type means:
# "explain gene expression by donor identity AND product type"
design <- model.matrix(~ donor + product_type, data = samplesheet)

# Estimate the correlation between the two wells from the same donor
# This accounts for the fact that 2 wells per donor are not independent
corfit <- duplicateCorrelation(logcpm, design, block = samplesheet$donor)

# Print the inter-donor correlation
# Values between 0.1-0.5 are typical for biological replicates
cat("Inter-well correlation:", corfit$consensus.correlation, "\n")

# Apply voom transformation
# This estimates the mean-variance relationship and assigns
# precision weights to each observation
v <- voom(y_filt, design,
          block = samplesheet$donor,
          correlation = corfit$consensus.correlation,
          plot = TRUE)
```

# DEG analysis
```{r deg_analysis, include=TRUE}
# Fit the linear model to the voom-transformed data
# This estimates the coefficients for each gene
fit <- lmFit(v, design,
             block = samplesheet$donor,
             correlation = corfit$consensus.correlation)

# Apply empirical Bayes moderation
# This borrows information across genes to get more stable variance estimates
fit <- eBayes(fit)

# Extract DEGs for each comparison vs control
# adjust.method="BH" is Benjamini-Hochberg correction for multiple testing
results_distillate <- topTable(fit,
                                coef = "product_typedistillate",
                                number = Inf,
                                adjust.method = "BH",
                                sort.by = "P")

results_disposable <- topTable(fit,
                                coef = "product_typedisposable",
                                number = Inf,
                                adjust.method = "BH",
                                sort.by = "P")

results_juice <- topTable(fit,
                           coef = "product_typejuice",
                           number = Inf,
                           adjust.method = "BH",
                           sort.by = "P")

# How many significant DEGs in each comparison?
cat("Significant DEGs - distillate vs control:",
    sum(results_distillate$adj.P.Val < 0.05 & abs(results_distillate$logFC) > 1), "\n")
cat("Significant DEGs - disposable vs control:",
    sum(results_disposable$adj.P.Val < 0.05 & abs(results_disposable$logFC) > 1), "\n")
cat("Significant DEGs - juice vs control:",
    sum(results_juice$adj.P.Val < 0.05 & abs(results_juice$logFC) > 1), "\n")
```

# PCA
```{r pca, include=TRUE, fig.width=8, fig.height=6}
# Perform PCA on log2 CPM values to explore sample clustering
# t() transposes the matrix so samples are rows and genes are columns
pca <- prcomp(t(logcpm), scale. = FALSE)

# Calculate % variance explained by each PC
pca_var <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

# Create data frame for plotting
pca_df <- data.frame(
    PC1 = pca$x[, 1],
    PC2 = pca$x[, 2],
    sample_id = rownames(pca$x)
)

# Add treatment and donor information for coloring/shaping
pca_df$product_type <- samplesheet$product_type[match(pca_df$sample_id,
                                                        samplesheet$sample_id)]
pca_df$donor <- samplesheet$donor[match(pca_df$sample_id,
                                         samplesheet$sample_id)]

# Plot PCA coloured by product type, shaped by donor
ggplot(pca_df, aes(x = PC1, y = PC2, color = product_type, shape = donor)) +
    geom_point(size = 3) +
    theme_bw() +
    labs(title = "PCA of log2 CPM counts",
         x = paste0("PC1 (", pca_var[1], "%)"),
         y = paste0("PC2 (", pca_var[2], "%)"),
         color = "Product Type",
         shape = "Donor")
```

# Heatmap of most variable genes
```{r heatmap_variable, include=TRUE, fig.width=10, fig.height=8}
# Calculate variance for each gene across all samples
gene_var <- apply(logcpm, 1, var)

# Select top 500 most variable genes
top500 <- names(sort(gene_var, decreasing = TRUE))[1:500]
mat <- logcpm[top500, ]

# Create annotation data frame for samples
# This adds colored bars at the top of the heatmap showing treatment and donor
ann_col <- data.frame(
    product_type = samplesheet$product_type,
    donor = samplesheet$donor,
    row.names = samplesheet$sample_id
)

# Plot heatmap
pheatmap::pheatmap(
    mat,
    scale = "row",          # scale each gene to mean=0, sd=1 so colors show relative expression
    show_rownames = FALSE,  # too many genes to show names
    annotation_col = ann_col,
    main = "Top 500 most variable genes",
    cluster_cols = TRUE,    # cluster samples by expression similarity
    cluster_rows = TRUE     # cluster genes by expression similarity
)
```

# Sample distance heatmap
```{r sample_distance, include=TRUE, fig.width=8, fig.height=7}
# Calculate Euclidean distance between all pairs of samples
# t() transposes so we calculate distances between samples not genes
sampleDists <- dist(t(logcpm))
sampleDistMatrix <- as.matrix(sampleDists)

# Label rows with sample ID and treatment for readability
rownames(sampleDistMatrix) <- paste(samplesheet$sample_id,
                                     samplesheet$product_type,
                                     sep = "-")
colnames(sampleDistMatrix) <- NULL

# Blue color palette - darker = more similar, lighter = more different
colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

# Plot heatmap
pheatmap(
    sampleDistMatrix,
    clustering_distance_rows = sampleDists,
    clustering_distance_cols = sampleDists,
    col = colors,
    main = "Sample distance matrix"
)
```

# Volcano plots
```{r volcano, include=TRUE, fig.width=10, fig.height=8}
# Function to create a volcano plot for any comparison
plot_volcano <- function(results, title) {

    # Add significance labels
    results$significance <- "Not significant"
    results$significance[results$adj.P.Val < 0.05 & results$logFC > 1] <- "Upregulated"
    results$significance[results$adj.P.Val < 0.05 & results$logFC < -1] <- "Downregulated"

    # Get top 10 genes by adjusted p-value for labeling
    top10 <- head(results[order(results$adj.P.Val), ], 10)

    ggplot(results, aes(x = logFC, y = -log10(adj.P.Val), color = significance)) +
        geom_point(alpha = 0.5, size = 0.8) +
        scale_color_manual(values = c("Upregulated" = "red",
                                      "Downregulated" = "blue",
                                      "Not significant" = "grey")) +
        # Add vertical lines at fold change thresholds
        geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
        # Add horizontal line at p-value threshold
        geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
        # Label top 10 genes
        ggrepel::geom_text_repel(data = top10, aes(label = rownames(top10)),
                                  size = 3) +
        theme_bw() +
        labs(title = title,
             x = "Log2 Fold Change",
             y = "-log10(adjusted p-value)",
             color = "Significance")
}

# Plot volcano for each comparison
plot_volcano(results_distillate, "Distillate vs Control")
plot_volcano(results_disposable, "Disposable vs Control")
plot_volcano(results_juice, "Juice vs Control")
```

# Top DEG expression plots
```{r top_degs, include=TRUE, fig.width=12, fig.height=8}
# Function to plot expression of top N genes for any comparison
plot_top_genes <- function(results, title, n = 9) {

    # Get top N genes by adjusted p-value
    top_genes <- rownames(head(results[order(results$adj.P.Val), ], n))

    # Extract their log2 CPM values
    top_expr <- as.data.frame(t(logcpm[top_genes, ]))
    top_expr$sample_id <- rownames(top_expr)

    # Add treatment information
    top_expr$product_type <- samplesheet$product_type[match(top_expr$sample_id,
                                                             samplesheet$sample_id)]

    # Melt to long format for ggplot
    melted <- reshape2::melt(top_expr,
                              id.vars = c("sample_id", "product_type"),
                              variable.name = "gene",
                              value.name = "logCPM")

    # Plot expression per gene coloured by treatment
    ggplot(melted, aes(x = product_type, y = logCPM, color = product_type)) +
        geom_jitter(width = 0.2, size = 2) +
        stat_summary(fun = mean, geom = "crossbar", width = 0.5) +
        facet_wrap(~ gene, scales = "free_y", ncol = 3) +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              legend.position = "none") +
        labs(title = title,
             x = "Product Type",
             y = "Log2 CPM")
}

# Plot top genes for each comparison
plot_top_genes(results_distillate, "Top DEGs - Distillate vs Control")
plot_top_genes(results_disposable, "Top DEGs - Disposable vs Control")
plot_top_genes(results_juice, "Top DEGs - Juice vs Control")
```

# GSEA pathway analysis
```{r gsea, include=TRUE}
# Load human gene sets from MSigDB - using Hallmarks collection
# Hallmarks are 50 well-defined biological processes
msig_human <- msigdbr::msigdbr(species = "Homo sapiens", category = "H") %>%
    dplyr::select(gs_name, ensembl_gene) %>%
    dplyr::filter(!is.na(ensembl_gene), ensembl_gene != "") %>%
    dplyr::distinct()

# Convert to list format required by fgsea
pathways <- split(msig_human$ensembl_gene, msig_human$gs_name)

# Function to run GSEA for any comparison
run_gsea <- function(results, title) {

    # Rank genes by log2 fold change
    # GSEA uses this ranking to find pathways enriched at top or bottom
    ranks <- results$logFC
    names(ranks) <- rownames(results)
    ranks <- sort(ranks, decreasing = TRUE)

    # Run GSEA
    set.seed(1)
    fg <- fgsea(
        pathways = pathways,
        stats = ranks,
        minSize = 15,   # minimum genes in pathway
        maxSize = 500   # maximum genes in pathway
    ) %>% as_tibble() %>% arrange(padj)

    # Print top results
    cat("\nTop pathways -", title, "\n")
    print(head(fg[, c("pathway", "NES", "pval", "padj")], 10))

    return(fg)
}

# Run GSEA for each comparison
gsea_distillate <- run_gsea(results_distillate, "Distillate vs Control")
gsea_disposable <- run_gsea(results_disposable, "Disposable vs Control")
gsea_juice <- run_gsea(results_juice, "Juice vs Control")
```

# GSEA plots
```{r gsea_plots, include=TRUE, fig.width=12, fig.height=8}
# Function to plot top enriched pathways for any comparison
plot_gsea <- function(gsea_results, title, n = 10) {

    # Select top n positive and top n negative NES pathways
    top_pos <- gsea_results %>%
        filter(NES > 0, padj < 0.05) %>%
        arrange(desc(NES)) %>%
        head(n)

    top_neg <- gsea_results %>%
        filter(NES < 0, padj < 0.05) %>%
        arrange(NES) %>%
        head(n)

    top_pathways <- bind_rows(top_pos, top_neg) %>%
        arrange(NES)

    # Check if any significant pathways found
    if(nrow(top_pathways) == 0) {
        cat("No significant pathways found for", title, "\n")
        return(NULL)
    }

    # Dotplot of top pathways
    ggplot(top_pathways, aes(x = NES,
                              y = reorder(pathway, NES),
                              size = -log10(padj),
                              color = NES > 0)) +
        geom_point() +
        scale_color_manual(values = c("TRUE" = "red", "FALSE" = "blue"),
                           labels = c("TRUE" = "Upregulated",
                                      "FALSE" = "Downregulated")) +
        labs(title = title,
             x = "Normalized Enrichment Score (NES)",
             y = "Pathway",
             size = "-log10(padj)",
             color = "") +
        theme_bw() +
        theme(axis.text.y = element_text(size = 8))
}

# Plot top pathways for each comparison
plot_gsea(gsea_distillate, "Top Pathways - Distillate vs Control")
plot_gsea(gsea_disposable, "Top Pathways - Disposable vs Control")
plot_gsea(gsea_juice, "Top Pathways - Juice vs Control")
```


# Save results
```{r save_results, echo=FALSE}
out_dir <- dirname(snakemake@output[[1]])
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# Save DEG results
write.csv(
  results_distillate,
  file.path(out_dir, "DEG_distillate_vs_control.csv")
)
write.csv(
  results_disposable,
  file.path(out_dir, "DEG_disposable_vs_control.csv")
)
write.csv(
  results_juice,
  file.path(out_dir, "DEG_juice_vs_control.csv")
)

# Helper to flatten list columns for CSV export
flatten_for_csv <- function(df) {
  df <- as.data.frame(df)
  is_list_col <- vapply(df, is.list, logical(1))

  if (any(is_list_col)) {
    df[is_list_col] <- lapply(df[is_list_col], function(col) {
      vapply(col, function(x) {
        if (is.null(x) || length(x) == 0) {
          NA_character_
        } else {
          paste(as.character(x), collapse = ";")
        }
      }, character(1))
    })
  }

  df
}

# Save GSEA results
write.csv(
  flatten_for_csv(gsea_distillate),
  file.path(out_dir, "GSEA_distillate_vs_control.csv"),
  row.names = FALSE
)
write.csv(
  flatten_for_csv(gsea_disposable),
  file.path(out_dir, "GSEA_disposable_vs_control.csv"),
  row.names = FALSE
)
write.csv(
  flatten_for_csv(gsea_juice),
  file.path(out_dir, "GSEA_juice_vs_control.csv"),
  row.names = FALSE
)
